---
title: "SWDIS raw analysis"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(choroplethr)
library(choroplethrMaps)
library(noncensus)
```

# Cleaning up the water systems

Get an overview of the variables.

```{r}
read_csv("data/swdis/WATER_SYSTEM.csv", n_max=10000) %>%
  summary()
```

```{r}
water_system <- read_csv("data/swdis/WATER_SYSTEM.csv", 
                         col_types=cols_only(PWSID=col_character(),
                                             POPULATION_SERVED_COUNT=col_integer(),
                                             ORG_NAME=col_character(),
                                             PWS_TYPE_CODE=col_factor(),
                                             ZIP_CODE=col_character()
                         ))
summary(water_system)

head(water_system)
```

Cleanup some of the data.

```{r}
clean_water_system <- water_system %>%
  na.omit() %>%
  filter(POPULATION_SERVED_COUNT<1000000, POPULATION_SERVED_COUNT>500, PWS_TYPE_CODE=="CWS")
summary(clean_water_system)
```

There are still population sized outliers.

Let's look at how many people are served.

```{r}
clean_water_system %>% 
  ggplot(aes(x=POPULATION_SERVED_COUNT)) +
  geom_histogram(bins=100)
```

Let's trim off the outliers
```{r}
clean_water_system %>% 
  filter(POPULATION_SERVED_COUNT<10000) %>%
  ggplot(aes(x=POPULATION_SERVED_COUNT)) +
  geom_histogram(bins=100)
```

# Cleanup the violations

```{r}
read_csv("data/swdis/VIOLATION.csv", n_max=10000) %>% head()
```

We want to look at:
* CONTAMINANT_CODE
* PWSID
* COMPL_PER_BEGIN_DATE
* COMPL_PER_END_DATE

```{r}
violations <- read_csv("data/swdis/VIOLATION.csv",
                       col_types = cols_only(
  CONTAMINANT_CODE=col_character(),
  PWSID=col_character(),
  COMPL_PER_BEGIN_DATE=col_date(format="%d-%b-%y"),
  COMPL_PER_END_DATE=col_date(format="%d-%b-%y")
))
head(violations)
```

Get the contaminant codes.

```{r}
contaminant_codes <- read_csv("data/swdis/contaminant-codes.csv") %>%
  rename(CONTAMINANT_CODE=CODE, CONTAMINANT_NAME=NAME, CONTAMINANT_TYPE_CODE=TYPE_CODE)
head(contaminant_codes)

contaminant_group_codes <- read_csv("data/swdis/contaminant-group-codes.csv") %>%
  rename(CONTAMINANT_CODE=`Analyte Code`, CONTAMINANT_NAME=`Analyte Name`, CONTAMINANT_GROUP=`Group Name`, CONTAMINANT_GROUP_CODE=`Group Code`)
contaminant_group_codes %>% head()
```

Join this with our cleaned up PWS ids.

```{r}
clean_violations <- violations %>%
  inner_join(clean_water_system,  by="PWSID") %>%
  inner_join(contaminant_group_codes, by="CONTAMINANT_CODE")
nrow(clean_violations)
head(clean_violations)
```

Let's do some count of violation types.

```{r}
tallied_violations <- clean_violations %>%
  group_by(CONTAMINANT_NAME) %>%
  summarize(n=n()) %>%
  arrange(desc(n))

tallied_violations
```

```{r}
clean_violations %>%
  group_by(CONTAMINANT_GROUP) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) %>%
  head(n=20)
```

Get FIPS region codes.

```{r}
library(choroplethr)
data(zip_codes)
data(counties)

violations_by_region <- clean_violations %>%
  inner_join(zip_codes, by=c("ZIP_CODE"="zip")) %>%
  rename(region=fips) %>%
  group_by(region) %>%
  summarize(count_violations=n(), total_pop=sum(POPULATION_SERVED_COUNT))
head(violations_by_region)
```

```{r hold=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
water_system_fips <- clean_water_system %>%
  inner_join(zip_codes, by=c("ZIP_CODE"="zip")) %>%
  rename(region=fips) %>%
  group_by(region) %>%
  summarize(total_pop=sum(POPULATION_SERVED_COUNT))

county_choropleth(water_system_fips %>% rename(value=total_pop), state_zoom=c("massachusetts"))
```

```{r hold=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
county_choropleth(violations_by_region %>% rename(value=total_pop), state_zoom=c("massachusetts"))
```

Why are there no violations listed for Boston itself?


```{r hold=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
county_choropleth(violations_by_region %>% rename(value=count_violations), state_zoom=c("massachusetts"))
```

Why is suffolk county missing? Let's look at the boston zip code.

```{r}
clean_water_system %>%
  filter(ZIP_CODE=="02155" | ZIP_CODE=="02122")

water_system %>%
  filter(ZIP_CODE=="02155" | ZIP_CODE=="02122")

water_system %>%
  filter(grepl("BOSTON",ORG_NAME))
```

Try to convert county fips code to a categorical data to get the county name.

```{r}
counties2 <- counties %>%
  mutate(fips = as.integer(paste0(state_fips, county_fips))) %>%
  select(fips, county_name)
counties2 %>% head()
```

Lookup FIPS code for the boston facilities in the `zip_codes` dataframe.

```{r}
boston_systems <- water_system %>%
  filter(grepl("BOSTON", ORG_NAME))
boston_systems <- boston_systems %>%
  left_join(zip_codes, by=c("ZIP_CODE" = "zip")) %>%
  left_join(counties2, by="fips") %>%
  select(ORG_NAME,  county_name, POPULATION_SERVED_COUNT,ZIP_CODE, fips, PWSID)
boston_systems
```

So interestingly here we do have Suffolk county showing up correctly. So it seems that there are no violations in Suffolk county itself.

Let's first plot the populations for the systems that have BOSTON in their name in MA.

```{r}
boston_systems %>%
  rename(value=POPULATION_SERVED_COUNT, region=fips) %>% 
  select(value, region) %>% 
  na.omit() %>%
  group_by(region) %>% summarize(value=sum(value)) %>%
  county_choropleth(state_zoom=c("massachusetts"))
```

Let's look if we can find violations for the main boston waterstation.

```{r}
boston_pwsid <- "MA3035000"
boston_systems %>% filter(PWSID==boston_pwsid)
```

```{r}
violations %>%
  filter(PWSID == boston_pwsid)
```

What is a bit annoying is that the EWG website lists quite a few violations for the MWRA on its website: [EWG for MWRA](https://www.ewg.org/tapwater/system.php?pws=MA3035000). Their source data comes from the EPA ECHO database, but [that database](https://echo.epa.gov/detailed-facility-report?fid=MA3035000&sys=SDWIS) doesn't show any violations either. I wonder if this is due to the date range, the EWG website mentiones 2010-2015, while the ECHO site shows data for the last 5 years. The [SWDIS search](https://ofmpub.epa.gov/enviro/sdw_report_v3.first_table?pws_id=MA3035000&state=MA&source=Surface%20water%20purchased&population=617594&sys_num=0) also doesn't show up any violations.

Let's see if we get to this data by taking all the facilities labeled MWRA into account.

```{r}
mwra_systems <- water_system %>%
  filter(grepl("MWRA", ORG_NAME))
mwra_systems <- mwra_systems %>%
  left_join(zip_codes, by=c("ZIP_CODE" = "zip")) %>%
  left_join(counties2, by="fips") %>%
  select(ORG_NAME,  county_name, POPULATION_SERVED_COUNT, PWSID)
mwra_systems
```

```{r}
mwra_pwsids <- mwra_systems$PWSID
violations %>%
  right_join(mwra_systems, by="PWSID") %>%
  group_by(PWSID,  CONTAMINANT_CODE) %>%
  summarize(n=n()) %>%
  arrange(PWSID, desc(n)) %>%
  left_join(contaminant_codes, by="CONTAMINANT_CODE") %>%
  left_join(mwra_systems, by="PWSID") %>%
  select(PWSID, ORG_NAME, POPULATION_SERVED_COUNT, n, CONTAMINANT_NAME) %>%
  arrange(desc(POPULATION_SERVED_COUNT), desc(n))
```

The EWG database shows Bromodichloromethane as one of the contaminants. Do we even know about this?

```{r}
contaminant_codes %>%
  filter(grepl(".*ROMODI.*", CONTAMINANT_NAME))
```

Let's look for bromodichloromethane violations in Massachusetts.

```{r}
violations %>%
  filter(CONTAMINANT_CODE==2943) %>%
  group_by(PWSID) %>%
  summarize(n=n()) %>% 
  left_join(water_system, by="PWSID") %>%
  select(PWSID, ORG_NAME, POPULATION_SERVED_COUNT, ZIP_CODE, n) %>%
  left_join(zip_codes, by=c("ZIP_CODE" = "zip"))  %>%
  filter(state == "MA") %>%
  arrange(desc(POPULATION_SERVED_COUNT))
```

sdfsdf
